---
title: "Data Visualization Project"
author: "Barthelemy Giordana"
date: "6/19/2018"
output: html_document
---

Installing (if needed) and loading packages for the project.

```{r setup, include=FALSE}
## install packages if necessary
list.of.packages <- c("janitor", "ggplot2", "dplyr", "lubridate", "zoo", "forcats", "ggthemes", "knitr", "rmarkdown")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)

## load packages
library(janitor)
library(ggplot2)
library(dplyr)
library(lubridate)
library(zoo)
library(forcats)
library(ggthemes)
library(knitr)

## set working directory
knitr::opts_knit$set(root.dir = '/cloud/project')
```

# Part 1: Exploratory Data Analysis

The data we'll be using for this part of the project were downloaded from [kaggle](https://www.kaggle.com/aaronschlegel/seattle-pet-licenses/data), and include information about "pet licenses issued by the Seattle Animal Shelter between 2005 and early 2017." We'll be exploring these data and generating a few exploratory plots in the first part of the project.


## The Data

First, we'll read the data in from our `data/raw_data` directory.

```{r pet-data, warning=FALSE} 
setwd("dataviz_project")
pets <- read.csv("data/raw_data/seattle_pet_licenses.csv", stringsAsFactors = FALSE)
```

## Explore the Data

Exploring the data.

```{r pet-explore, warning=FALSE} 
## Include code here to explore the data 
## and answer the questions in the exercise
## number of licenses issued
nrow(pets)
## how many pet names
n_distinct(pets$animal_s_name)
## how many species
n_distinct(pets$species)
## name of the species
count(pets,species)
```

## Visualize the Data

```{r pet-barchart, warning=FALSE} 
## visual breakdown of how many of each species 
## are included in the dataset
## barchart
ggplot(data = pets) +
  geom_bar(aes(x=species))
```

```{r pet-cat-table, warning=FALSE} 
## Table: Most frequent Cat Name 
pets %>% 
  filter(species == "Cat", animal_s_name != "") %>%
  group_by(animal_s_name) %>% 
  summarise(n = n()) %>% 
  arrange(-n) %>%
  top_n(n = 10) %>%
  knitr::kable(., caption = "Top 10 Cat Names in Seattle")
```

```{r pet-dog-table, warning=FALSE} 
## Table: Most frequent Dog Name
pets %>% 
  filter(species == "Dog", animal_s_name != "") %>%
  group_by(animal_s_name) %>% 
  summarise(n = n()) %>% 
  arrange(-n) %>%
  top_n(n = 10) %>%
  knitr::kable(., caption = "Top 10 Dog Names in Seattle")
```

```{r pet-linechart, warning=FALSE} 
## add date and ym columns
pets$date <- lubridate::ymd_hms(pets$license_issue_date)
pets$ym <- as.yearmon(pets$date, "%y%m")

## how the number of licenses recorded has changed over time
pets %>%
  ## group by yearmonth (`ym`)
  group_by(ym) %>% 
  ## count number within each group
  summarize(n=n()) %>%
  ggplot(., aes(ym, n)) + 
  ## geom name for line chart
  geom_line() +
  scale_x_yearmon() + 
  xlab("") + 
  ylab("Number of licenses")
```

```{r pet-plot, warning=FALSE} 
### Explore the data -- create your own plot!
pets %>%
  ## group by yearmonth (`ym` , `species`)
  group_by(ym,species) %>% 
  ## count number within each group
  summarize(n=n()) %>%
  ggplot(., aes(x = ym,y = n, color = species)) + 
  ## geom name for line chart
  geom_line() +
  scale_x_yearmon() + 
  xlab("") + 
  ylab("Number of licenses per species")
```

# Part 2: Explanatory Data Analysis

The data used in this part of the project were dowloaded from [FiveThirtyEight](https://data.fivethirtyeight.com/) - `steak-survey`.  They were originally used in the article: [How Americans Like Their Steak](https://fivethirtyeight.com/features/how-americans-like-their-steak/). The goal of this part of the project will be to recreate the data visualization used in this article.

## The Data

```{r meat-data, warning=FALSE} 
## read in the data
  setwd("dataviz_project")
  steak <- read.csv("data/raw_data/steak-risk-survey.csv", stringsAsFactors = FALSE) %>%
  ## remove first row which just includes the word "Response" in each column
  slice(2:n())
```

## Explore the Data

```{r meat-explore, warning=FALSE} 
## Include code here to explore the data
nrow(steak)
## and answer the questions in the exercise
## number of steak eater
steak %>% filter(Do.you.eat.steak.=="Yes") %>% nrow()
## number of steak wellness
steak %>% filter(How.do.you.like.your.steak.prepared. != "") %>% select(How.do.you.like.your.steak.prepared.) %>% n_distinct()

```

## Wrangle the Data

```{r meat-wrangle, warning=FALSE} 
## wrangle the steak data set
pref <- steak %>% 
  mutate(steak_pref = factor(How.do.you.like.your.steak.prepared.,
                             levels = c("Well",
                                        "Medium Well",
                                        "Medium",
                                        "Medium rare",
                                        "Rare"))) %>% 
  filter(steak_pref != "") %>% 
  group_by(steak_pref) %>% 
  summarise(n = n()) %>%
  mutate(prop = n / sum(n))
```

## Visualize the Data

```{r meat-visualize, warning=FALSE}
## generate the plot 
p <- ggplot(pref) +
  ## bar chart
  geom_bar(aes(x = steak_pref, y = prop, fill = steak_pref),
                stat = 'identity', 
                width = 0.7) + 
  ## this adds text labels (you don't have to change anything here)
  geom_text(aes(label = paste0(as.integer(prop*100),"%"),
                x = steak_pref,
                y = prop), 
            stat= "identity", 
            hjust = -0.2, 
            size = 5,
            color = "grey40") +
  ## flip coordinates to make horizontal box plot
  coord_flip() +
  ## change the colors of the bars
  scale_fill_manual(values = c("#370B01", 
                           "#70402C",
                           "#955732", 
                           "#B36153",
                           "#B5413F")) +
  ## change the scale/labels of the steak-wellness variable (x-axis)
  scale_x_discrete(limits = levels(fct_rev(pref$steak_pref)),
                labels = c("Well", 
                           "Medium Well" = "Medium-well",
                           "Medium",
                           "Medium rare"="Medium-rare",
                           "Rare")) +
  ## change the scale/lables of the percent axis (y-axis)
  scale_y_continuous(labels = scales::percent, 
                expand = c(mult = c(0,0), 
                           add = c(0,0.035))) +
  ## change the title, subtitle, and caption
  labs(title="'How Do You Like Your Steak Prepared?'", 
                subtitle="From a survey of 432 steak-eating Americans",
                caption="FiveThirtyEight: Surveymonkey") +
  ## change the theme (use ggthemes)
  theme_void() +
  ## fine tune the theme
  theme(axis.text = element_text(size = 14), 
                title = element_text(size = 16), 
                legend.position="none", 
                plot.caption=element_text(size = 12), 
                panel.grid.major = element_blank(), 
                panel.grid.minor = element_blank(),
                axis.text.x = element_blank() 
  )

p
```

## Save the Plot

```{r save-plot, warning=FALSE}
## save plot to figures/explanatory_figures directory
setwd("dataviz_project")
ggsave(plot = p, filename = "figures/explanatory_figures/steak_R.png", width = 8, height = 4)
```

## Visualize the Data

```{r meat-plot, warning=FALSE} 
### Explore the data -- create your own plot!
## number of valid respondent
steak %>% filter(Gender != "" & Do.you.ever.smoke.cigarettes. != "")  %>% nrow()
## number of smokers/non smokers
steak %>% filter(Gender != "" & Do.you.ever.smoke.cigarettes. == "Yes")  %>% nrow()
steak %>% filter(Gender != "" & Do.you.ever.smoke.cigarettes. == "No")  %>% nrow()

g <- steak %>% filter(Gender != "" & Do.you.ever.smoke.cigarettes. != "")  %>%              ggplot()+geom_bar(aes(x=Gender,fill=Do.you.ever.smoke.cigarettes.)) +                     scale_fill_manual(labels=c("Non smokers","Smokers"),values=c("lightblue","grey40")) +
  ylab("Number of people per gender") +
  theme_classic() +
  labs(title="Number of smokers/non smokers per gender", 
  subtitle="From a survey of 510 Americans",
  caption="Source : LONGHORN STEAKHOUSE") +
  theme(
    legend.title = element_blank()
  )

g
```

## Save this plot 

```{r save-meat-plot, warning=FALSE}
## save plot to figures/explanatory_figures directory
setwd("dataviz_project")
ggsave(plot = g, filename = "figures/explanatory_figures/smokers_R.png")
```
